Division and Factorization of p-Adic Polynomials

### 2016-01-28 polynomials/muldivfact.tex
Small progress toward num. results
More interesting exemples.
Need a better understanding of the computation, and then would be ready
to give an array of results.

\todo{Write introduction.}

@@

@@

@@

\section{Precision data}

@@

\section{Euclidean division}

@@

\section{Slope factorization}

@@

\begin{prop}
\label{prop:precA1}
The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
and its differential at at some point $P$ is the linear mapping:
$$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
where $(A^{(1)}, B^{(1)}) = F(P)$ and $V^{(1)}$ is the inverse of 
$B^{(1)}$ modulo $A^{(1)}$.
\end{prop}

>>>>

Polynomial factorization is a fundamental problem in computational algebra.
The algorithms used to solve it depend on the ring of coefficients, with finite fields,
local fields, number fields and rings of integers of particular interest to number theorists.
In this article, we focus on a task that forms a building block for factorization
algorithms over complete discrete valuation fields: the decomposition into factors based on the
slopes of the Newton polygon.

The Newton polygon of a polynomial $f(X) = \sum a_i X^i$ over such a field is given by
the convex hull of the points $(i, \val(a_i))$ and the point $(0, +\infty)$.  The lower
boundary of this polygon consists of line segments $(x_j,y_j)$ -- $(x_{j+1},y_{j+1})$
of slope $s_j$.  The slope factorization of $f(X)$ expresses $f(X)$ as a product
of polynomials $g_j(X)$ with degree $x_{j+1} - x_j$ whose roots all have valuation $-s_j$.
Our main result is a new algorithm for computing these $g_j(X)$.

Any computation with $p$-adic fields must work with approximations modulo finite
powers of $p$, and one of the key requirements in designing an algorithm is
an analysis of how the precision of the variables evolve over the computation.
We work with precision models developed by the same authors
\cite{caruso-roe-vaccon:14a}*{\S 4.2}, focusing on the lattice and Newton models.
As part of the analysis of the slope factorization algorithm, we describe
how the precision of the quotient and remainder depend on the input polynomials
in Euclidean division..

@@

Suppose that the Newton polygon of $P(X) = \sum_{i=0}^n a_i X^i$
has a break at $i=d$.  Set $A_0 = \sum_{i=0}^d a_i X^i$, $V_0 = 1$ and
\begin{align*}
A_{i+1} &= A_i + (V_i P \% A_i) \\
B_{i+1} &= P // A_{i+1} \\
V_{i+1} &= (2V_i - V_i^2B_{i+1}) \% A_{i+1}.
\end{align*}
Our main result is Theorem \ref{theo:slope-factor},
which states that the sequence $(A_i)$ converges
to a divisor of $P$.

@@

After setting notation, in
Section \ref{sec:prec_data} we recall various models for tracking
precision in polynomial arithmetic.  We give some background
on Newton polygons and explain how using lattices to store
precision can allow for extra \emph{diffuse} $p$-adic digits that
are not localized on any single coefficient.

In Section \ref{sec:quo_rem}, we consider Euclidean division.
We describe in Theorem \ref{theo:EDivisionNP}
how the Newton polygons of the quotient and remainder
depend on numerator and denominator.  We use this
result to describe in Proposition \ref{prop:NewtonprecEuclide}
the precision evolution in Euclidean division using
the Newton precision model.  We then compare the
precision performance of Euclidean division in the
jagged, Newton and lattice models experimentally,
finding different behavior depending on the modulus.

Finally, in Section \ref{sec:slope_fac} we describe
our slope factorization algorithm, which is based on
a Newton iteration.  Unlike other algorithms for
slope factorization, ours does not require working
in an extension field \todo{citation?}.  In Theorem
\ref{theo:slope-factor} we define a sequence
of polynomials that will converge to the factors
determined by an extremal point in the Newton
polygon.  We then discuss the precision behavior
of the algorithm.

@@

\section{Precision data} \label{sec:prec_data}

@@

\section{Euclidean division} \label{sec:quo_rem}

@@

\section{Slope factorization} \label{sec:slope_fac}

@@

\begin{prop}
\label{prop:precA1}
The application $F_A : P \mapsto A^{(1)}$ is of class $C^1$ on $P_\app + H_P$
and its differential at some point $P$ is the linear mapping:
$$dP \mapsto dA^{(1)} = (V^{(1)} \: dP) \mod A^{(1)}$$
where $(A^{(1)}, B^{(1)}) = F(P)$ and $V^{(1)}$ is the inverse of 
$B^{(1)}$ modulo $A^{(1)}$.
\end{prop}
